---
title: "Data wrangling with tidyr and dplyr"
author: "Eirini Zormpa"
institute: "The RSA"
title-slide-attributes:
    data-background-image: https://images.unsplash.com/photo-1692606944142-b0b4cde990e8?q=80&w=2080&auto=format&fit=crop&ixlib=rb-4.0.3&ixid=M3wxMjA3fDB8MHxwaG90by1wYWdlfHx8fGVufDB8fHx8fA%3D%3D
    data-background-opacity: "0.7"
format:
  revealjs:
    theme: [moon, custom.scss]
    slide-number: true
from: markdown+emoji
---

```{r}
#| echo: false

library(here)
library(readr)
library(countdown)

library(tidyr)
library(dplyr)

library(forcats)
library(lubridate)
```

```{r}
#| echo: false
#| eval: true

census_data <- read_csv(here("data_raw/synthetic-census-data.csv"))
```


## Last time you learned how to:

::: {data-id="box1" style="background: #03ecdd; width: 100px; height: 75px; margin: 10px; position: absolute; top: -100px; left: -10px"}

:::

- :white_check_mark: Read data into R
- :white_check_mark: Understand and manipulate `tibbles`
- :x: Understand and manipulate `factors`
- :x: Alternate between date formats

## Factors

::: {data-id="box1" style="background: #03ecdd; width: 100px; height: 75px; margin: 10px; position: absolute; top: -100px; left: -10px"}

:::

R has a special data class, called `factor`, to deal with *categorical data*. Factors:

- are stored as **integers** associated with labels, though they look like character vectors

. . .

- can be ordered (ordinal) or unordered (nominal)

. . .

-  create a structured relation between the different `levels` (values) of a categorical variable, such as days of the week or responses to a question in a survey


## Exercise

::: {data-id="box1" style="background: #03ecdd; width: 100px; height: 75px; margin: 10px; position: absolute; top: -100px; left: -10px"}

:::

:alarm_clock: **5 mins**

```{r}
countdown(minutes = 5)
```

1. In the `central_heating` variable, rename “no”, “yes”, and "unknown" to “No”, “Yes” and "Unknown" respectively.
2. Recreate the barplot such that “Unknown” is last.

## Solution

::: {data-id="box1" style="background: #03ecdd; width: 100px; height: 75px; margin: 10px; position: absolute; top: -100px; left: -10px"}

:::

```{r}
#| echo: true
#| eval: true

census_data$central_heating <- fct_recode(census_data$central_heating,
                              No = "no",
                              Unknown = "unknown",
                              Yes = "yes")

census_data$central_heating <- factor(census_data$central_heating, levels = c("No", "Yes", "Unknown"))

plot(census_data$central_heating)

```

# Data wrangling with dplyr and tidyr

::: {data-id="box1" style="background: #03ecdd; width: 100px; height: 75px; margin: 10px; position: absolute; top: -100px; left: -10px"}

:::

## Learning objectives

- Subset columns or rows with `select` or `filter` and create new columns with `mutate`.
- Link the output of one function to the input of another function with the ‘pipe’ operator `%>%`.
- Use `summarise`, `group_by`, and `count` to split a data frame into groups of observations, apply summary statistics for each group, and then combine the results.
- Export a dataframe to a .csv and .tsv file.

## Exercise 3.1

::: {data-id="box1" style="background: #03ecdd; width: 100px; height: 75px; margin: 10px; position: absolute; top: -100px; left: -10px"}

:::

:alarm_clock: **5 mins**

```{r}
countdown::countdown(minutes = 5)
```


Using pipes, subset `census_data` to include responses from participants based in London and retain only the columns `household_size`, `dwelling_type`, and `cars`


## Exercise 3.1 solution

::: {data-id="box1" style="background: #03ecdd; width: 100px; height: 75px; margin: 10px; position: absolute; top: -100px; left: -10px"}

:::

```{r}
#| echo: true
#| eval: false

census_data %>%
  filter(region == "London") %>%
  select(household_size, dwelling_type, cars)

```

. . .

Note that if you `select` before you `filter`, your code won't run.
That's because you're not retaining the variable that you use in your filtering.
When piping, **order matters**!


## Exercise 3.2

::: {data-id="box1" style="background: #03ecdd; width: 100px; height: 75px; margin: 10px; position: absolute; top: -100px; left: -10px"}

:::

:alarm_clock: **10 mins**

```{r}
countdown::countdown(minutes = 10)
```

1. How many respondents lived in each `dwelling_type`?
2. Use `group_by()` and `summarise()` to find the median, min, and max number of bedrooms for each `dwelling_type`. Also add the number of observations (hint: see `?n`)
3. What was the dwelling type with the most cars for each region?


## Exercise 3.2 solution

::: {data-id="box1" style="background: #03ecdd; width: 100px; height: 75px; margin: 10px; position: absolute; top: -100px; left: -10px"}

:::

```{r}
#| echo: true
#| eval: false

1.
census_data %>%
    count(dwelling_type)

2. 
census_data %>%
    group_by(dwelling_type) %>%
    summarise(
        median_bedrooms = median(bedrooms),
        min_bedrooms = min(bedrooms),
        max_bedrooms = max(bedrooms),
        n = n()
    )

3.
census_data %>%
  filter(!is.na(cars)) %>%
  group_by(region) %>%
  filter(cars == max(cars))

```

## Summary

::: {data-id="box1" style="background: #03ecdd; width: 100px; height: 75px; margin: 10px; position: absolute; top: -100px; left: -10px"}

:::

- :white_check_mark: Subset columns or rows with `select` or `filter` and create new columns with `mutate`.
- :white_check_mark: Link the output of one function to the input of another function with the ‘pipe’ operator `%>%`.
- :white_check_mark: Use `summarise`, `group_by`, and `count` to split a data frame into groups of observations, apply summary statistics for each group, and then combine the results.
- :white_check_mark: Export a dataframe to a `.csv` and `.tsv` file.